---
title: 'Topic 2: Exercise 1'
author: "Daniel Alonso"
date: "November 25th, 2020"
output: 'pdf_document'
---

#### Importing libraries

```{r, echo=TRUE, warning=FALSE, message=FALSE}
library(dplyr)
```

#### Importing data as described by exercise

```{r, echo=TRUE, warning=FALSE, message=FALSE}
d <- read.csv("../../datasets/Colleges.csv")
```

#### Replacing binary variable Private with 1 and 0

```{r, echo=TRUE, warning=FALSE, message=FALSE}
d$Private <- ifelse(d$Private == "Yes", 1, 0)
```

#### Selecting columns

```{r, echo=TRUE, warning=FALSE, message=FALSE}
d <- d %>% dplyr::select('Private','Apps','Accept','Enroll','F.Undergrad')
```

#### Calculating covariances

```{r, echo=TRUE, warning=FALSE, message=FALSE}
cov_matrix <- cov(d)
cov_matrix
```

\newpage

#### Calculating correlations

```{r, echo=TRUE, warning=FALSE, message=FALSE}
corr_matrix <- cov2cor(cov_matrix)
corr_matrix
```

## What information does the sample covariance provide?

We know that because the Private variable (binary variable) has only 2 possible values, its covariance with other variables is always going to be relatively small, and probably depends uniquely on the variance of that quantitative variable.

And we can prove this by simulating a covariance matrix:

```{r, echo=TRUE, warning=FALSE, message=FALSE}
size <- 1000
sim <- data.frame(matrix(rep(0,size*777),nrow=777))
names(sim) <- paste("sim",1:size,sep="_")
sim$Private <- d %>% dplyr::select('Private')
for (i in 1:size) {
    sim[i] <- rnorm(length(d$Private),mean=runif(1,min=1,max=100),sd=runif(1,min=1,max=5000))
}
Y <- cov(sim)[1,]
X <- diag(cov(sim))
plot(X,Y)
```

\newpage

## What information does the sample correlation provide?

### Scatter plot of our quantitative variables and the Private binary variable

```{r, echo=TRUE, warning=FALSE, message=FALSE}
plot(d$Private, d$Apps)
```