---
title: 'Topic 2: Exercise 1'
author: "Daniel Alonso"
date: "November 25th, 2020"
output: 'pdf_document'
---

#### Importing libraries

```{r, echo=TRUE, warning=FALSE, message=FALSE}
library(dplyr)
```

#### Importing data as described by exercise

```{r, echo=TRUE, warning=FALSE, message=FALSE}
d <- read.csv("../../datasets/Colleges.csv")
```

#### Replacing binary variable Private with 1 and 0

```{r, echo=TRUE, warning=FALSE, message=FALSE}
d$Private <- ifelse(d$Private == "Yes", 1, 0)
```

#### Selecting columns

```{r, echo=TRUE, warning=FALSE, message=FALSE}
data <- d %>% dplyr::select('Private','Apps','Accept','Enroll','F.Undergrad')
```

#### Calculating covariances

```{r, echo=TRUE, warning=FALSE, message=FALSE}
cov_matrix <- cov(data)
cov_matrix
```

\newpage

#### Calculating correlations

```{r, echo=TRUE, warning=FALSE, message=FALSE}
corr_matrix <- cov2cor(cov_matrix)
corr_matrix
```

## What information does the sample covariance provide?

We know that because the Private variable (binary variable) has only 2 possible values, its covariance with other variables is always going to be relatively small.

\newpage

## What information does the sample correlation provide?



### Scatter plot of our quantitative variables and the Private binary variable

```{r, echo=TRUE, warning=FALSE, message=FALSE}
plot(d$Private, d$Apps)
```